Supporting information

### 2018-09-07 second-reviewer-response/response_summary.txt

Dear Dr. David Ackerly,

Thank you for your detailed and thoughtful review of our manuscript. We have considered all of the comments provided by you and the two reviewers, and are pleased to present a revised version of our manuscript in response. I have included a point-by-point response to all comments in a separate document, but I will summarize the two largest changes here.

The primary concern expressed by you and reviewer 1 was the imputation of trait data in our analysis. We believe that these concerns stem primarily from a misunderstanding of our methods. Namely, unlike single imputation, where the imputation is a separate step in the analysis performed before the main parameter estimation, our manuscript uses multiple imputation, where imputation and parameter estimation happen simultaneously throughout the sampling procedure. We have clarified this distinction in the Methods, and included an extended demonstration of this technique in the Supporting Information (Method S1). In addition, we have provided data tables indicating the sample sizes and extent of imputation throughout our analysis as requested, but we did not perform the jackknifing analysis you requested because we did not feel it made sense in the context of multiple imputation as implemented in our manuscript.

An additional concern you expressed was the confounding effect of sampling---particularly, sample size and data range---on our correlation estimates. We agree that we did not give this sufficient attention in our initial draft, and we performed all of the analyses you suggested in this revision. These analyses show that although our results are not entirely independent of sampling artifacts, there is substantial variability in trait correlations that cannot be explained by sample size and/or data range, indicating that our results are in fact biologically relevant. We present these analyses in full in the supplementary information (Tables S3 and S4, Figures S1 and S2) and discuss them in the revised "Methods" and "Results" sections.

We hope that this revision is a step in the right direction in moving this manuscript towards publication at New Phytologist.

Best,
Alexey Shiklomanov

### 2018-09-07 manuscript/02_methods.Rmd
Previously unstaged changes from second revision

We emphasize that the imputation of missing data is performed iteratively as parameters are being estimated, such that each set of imputed values is conditioned on the current sampled state of the parameters.
This approach is therefore distinct from single imputation, where data are imputed first in a separate step prior to parameter estimation.
A demonstration of this multiple imputation approach and how it is used to estimate trait covariance is provided in a literate programming vignette in the supplementary information to this manuscript.

@@

To assess the impact of multivariate and hierarchical constraint on trait estimates,
we compared the mean and 95% confidence intervals of trait estimates for each PFT from each model (Fig.\ 2).
For reference, we also added the default parameter values of CLM 4.5 [Table 8.1 in @clm45_note] for SLA, $N_{mass}$, $N_{area}$, $V_{c,max,mass}$, and $V_{c,max,area}$ to Fig.\ 2.
To convert CLM's reported C:N ratio to $N_{mass}$, we assumed a uniform leaf C fraction of 0.46.
We then divided this calculated $N_{mass}$ by the reported SLA to obtain $N_{area}$.
We calculated $V_{c,max,mass}$ by multiplying the reported $V_{c,max,area}$ by the reported SLA.

>>>>

We emphasize that the imputation of missing data is performed iteratively as parameters are being estimated, such that each set of imputed values is conditioned on the current sampled state of the parameters.
This approach is therefore distinct from single imputation, where data are imputed first in a separate step prior to parameter estimation.
A demonstration of this multiple imputation approach and how it is used to estimate trait covariance is provided Supporting Information Method S1.

@@

To assess the impact of multivariate and hierarchical constraint on trait estimates,
we compared the mean and 95% confidence intervals of trait estimates for each PFT from each model (Fig.\ 2, Tab.\ S1 and S2).
For reference, we also added the default parameter values of CLM 4.5 [Table 8.1 in @clm45_note] for SLA, $N_{mass}$, $N_{area}$, $V_{c,max,mass}$, and $V_{c,max,area}$ to Fig.\ 2.
To convert CLM's reported C:N ratio to $N_{mass}$, we assumed a uniform leaf C fraction of 0.46.
We then divided this calculated $N_{mass}$ by the reported SLA to obtain $N_{area}$.
We calculated $V_{c,max,mass}$ by multiplying the reported $V_{c,max,area}$ by the reported SLA.

### 2018-09-07 manuscript/05_figures_tables.Rmd
Revise formatting for submission
Previously unstaged changes from second revision

# Tables {-}

@@

# Figure captions {-}

**Figure 1**: Sample sizes for each trait-PFT pair. $y$ axis is scaled logarithmically.

**Figure 2**: Mean and 95% confidence interval on best estimates of traits for each plant functional type from the univariate, multivariate, and hierarchical models.
For leaf lifespan and SLA, results were not significantly different between the mass- and area-based models, so only results from the mass-based model are shown.
For some PFT-trait combinations, where large error bars resulting from the relatively uninformative priors are substantially larger than the variability among means, the $y$ axes are constrained to facilitate comparison.

**Figure 3**: Relative uncertainty in PFT-level trait estimates as a function of sample size for each model type.
Lines represent linear models ($\log(y) = b_0 + b_1 \log(x)$) fit independently for each model type.
In general, differences in estimate uncertainty between the univariate and multivariate models were minimal at large sample sizes but increasingly important at low sample sizes.
However, differences in estimate uncertainty between the multivariate and hierarchical models were consistently negligible.

**Figure 4**: Pairwise trait mean and covariance estimates for all data pooled globally (black) and for each PFT (colored).
Covariance estimates not significantly different from zero ($p < 0.05$) are indicated by x symbols at the mean estimate.
$x$ and $y$ axes vary on a log scale, reflecting the fact that the model was fit using the base 10 log of all traits.
With the exception of leaf lifespan, pairwise covariances are consistent in direction but vary somewhat in magnitude between PFTs, and when comparing PFT-level and global estimates.
However, many pairwise covariances are not statistically significant, particularly (but not always) for undersampled traits and PFTs.

**Figure 5**: Mean and 95% CI on estimates of pairwise correlation coefficients
for all data pooled globally (dark grey) and for each PFT (colored).
For most PFT-trait pairs, correlations are mutually consistent in magnitude but vary in strength.

# References {-}

>>>>

<div custom-style="Heading, no number">Tables</div>

@@

### 2018-09-07 manuscript/index.Rmd
Previously unstaged changes from second revision

Section                   Word count          
--------                  --------------      ------------
Introduction              1060                Figures:
Materials and Methods     1695                5 (all colored)
Results                   1108
Discussion                1907                Tables:
Acknowledgements          88                  2
Total                     5858

>>>>

Section                   Word count
--------                  --------------      ------------
Introduction              1060                Figures: 5 (all colored)
Materials and Methods     1695                Tables: 2
Results                   1108                Supplementary figures: 2
Discussion                1907                Supplementary tables: 4
Acknowledgements          88                  Supplementary methods: 1
Total                     5858

### 2018-09-07 supplementary_information.Rmd
Previously unstaged changes from second revision

---
title: Supplementary information
author:
  - >
    Alexey N. Shiklomanov,
    Elizabeth M. Cowdery,
    Michael Bahn,
    Chaeho Byun,
    Joseph Craine,
    Andrés Gonzalez-Melo,
    Steven Jansen,
    Nathan Kraft,
    Koen Kramer,
    Vanessa Minden,
    Ülo Niinemets,
    Yusuke Onoda,
    Enio Egon Sosinski,
    Nadejda A. Soudzilovskaia,
    Michael C. Dietze
output: pdf_document
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \renewcommand{\thetable}{S\arabic{table}}
  - \renewcommand{\thefigure}{S\arabic{figure}}
---

@@

>>>>

---
title: Supporting information
author:
  - >
    Alexey N. Shiklomanov,
    Elizabeth M. Cowdery,
    Michael Bahn,
    Chaeho Byun,
    Joseph Craine,
    Andrés Gonzalez-Melo,
    Steven Jansen,
    Nathan Kraft,
    Koen Kramer,
    Vanessa Minden,
    Ülo Niinemets,
    Yusuke Onoda,
    Enio Egon Sosinski,
    Nadejda A. Soudzilovskaia,
    Michael C. Dietze
output: pdf_document
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \renewcommand{\thetable}{S\arabic{table}}
  - \renewcommand{\thefigure}{S\arabic{figure}}
---

@@

\clearpage

# Method S1: Estimating trait means and covariances through multiple imputation {-}

## Introduction {-}

Under _single_ imputation, the imputation step occurs once outside of the MCMC loop.
In _multiple_ imputation, as implemented in this paper, the imputation step is part of the MCMC loop,
such that imputed data values are conditioned on the current state of the parameters and vice-versa.
In the figure below,
$Y$ is the original data (with missing values),
$Y^*$ is the original data with missing values imputed,
$\mu_p$ and $\Sigma_p$ are values of the mean vector and covariance matrix (respectively) calculated only from the original data,
and $\mu^*_t$ and $\Sigma^*_t$ are the draws of the mean vector and covariance matrix (respectively) at MCMC iteration $t$.

```{tikz diagram, echo = FALSE}
\usetikzlibrary{positioning, calc, fit}
\begin{tikzpicture}[
	every node/.style = {shape = rectangle, draw = black},
	scale = 1.1
]
\node (Y) at (-1, 3) {Data $Y$};
\node (data params) at (-1, 2) {Calculate $\mu_p, \Sigma_p$};
\node (Ystar) at (-1, 1) {Impute $Y^* \mid \mu_p, \Sigma_p$};
\node (guess params) [align = center, right = of Ystar] {Initial guess\\$\mu^*_0$ $\Sigma^*_0$};
\node (mu star) at (0, 0) {Draw $\mu^*_t \mid Y^*, \Sigma^*_{t-1}$};
\node (sigma star) at (0, -1) {Draw $\Sigma^*_t \mid Y^*, \mu^*_t$};
\path [->] (Y) edge (data params);
\path [->] (data params) edge (Ystar);
\path [->] (Ystar) edge (mu star);
\path [->] (guess params) edge (mu star);
\path [->] (mu star) edge (sigma star);
\path [->, out = 0, in = 0] (sigma star.east) edge
	node (t plus one) [right, draw = none, font = {\footnotesize}] {$t = t + 1$}
	(mu star.east);
\node (mcmc loop) [fit = (mu star)(sigma star)(t plus one), shape = rectangle, draw = black!50] {};
\node (mcmc loop text) [align = center, anchor = west, draw = none] at (mcmc loop.east) {MCMC\\loop};
\node (title) [anchor = north, draw = none] at (mcmc loop.south) {Single imputation};
\end{tikzpicture}
\hspace{1cm}
\begin{tikzpicture}[
	every node/.style = {shape = rectangle, draw = black},
	scale = 1.1
]
\node (Y) at (-1, 1) {Data $Y$};
\node (guess params) [align = center] at (1, 1) {Initial guess\\$\mu^*_0$ $\Sigma^*_0$};
\node (Ystar) at (0, 0) {Impute $Y^*_t \mid Y, \mu^*_{t-1}, \Sigma^*_{t-1}$};
\node (mu star) at (0, -1) {Draw $\mu^*_t \mid Y^*_t, \Sigma^*_{t-1}$};
\node (sigma star) at (0, -2) {Draw $\Sigma^*_t \mid Y^*_t, \mu^*_t$};
\path [->] (Y) edge (Ystar);
\path [->] (guess params) edge (Ystar);
\path [->] (Ystar) edge (mu star);
\path [->] (mu star) edge (sigma star);
\path [->, out = 0, in = 0] (sigma star.east) edge
	node (t plus one) [right, draw = none, font = {\footnotesize}] {$t = t + 1$}
	(Ystar.east);
\node (mcmc loop) [fit = (Ystar)(mu star)(sigma star)(t plus one), shape = rectangle, draw = black!50] {};
\node (mcmc loop text) [align = center, anchor = west, draw = none] at (mcmc loop.east) {MCMC\\loop};
\node (title) [anchor = north, draw = none] at (mcmc loop.south) {Multiple imputation};
\end{tikzpicture}
```

## Demonstration {-}

Consider two positively correlated traits, A and B.

```{r create_traits}
set.seed(12345) # For reproducibility
library(mvtraits)

true_mu <- c(0, 0)
true_cov <- matrix(c(1, 0.8, 0.8, 1), nrow = 2)
Y_all <- random_mvnorm(1000, true_mu, true_cov)
colnames(Y_all) <- c("A", "B")
plot(Y_all, pch = 19)
```

Simulate missingness by randomly removing half of the data.

```{r remove_na}
Y <- Y_all
Y[sample.int(2000, 1000)] <- NA
plot(Y_all, pch = 19, col = "grey")
points(Y, pch = 19, col = "black")
legend(
  "topleft",
  legend = c("True missing", "Present"),
  col = c("grey", "black"),
  pch = 19,
  bg = "white"
)
```

Set an uninformative multivariate normal prior on $\mu$...
```{r prior_mu}
mu_0 <- c(0, 0)
Sigma_0 <- diag(10, 2)
```

...and an uninformative Wishart prior on $\Sigma$.
```{r prior_sigma}
v0 <- 2
S0 <- diag(10, 2)
```

Take an initial guess at the mean and covariance by drawing from their priors.

```{r initial}
mu_star <- random_mvnorm(1, mu_0, Sigma_0)[1,]
mu_star
Sigma_star <- rWishart(1, v0, S0)[,,1]
Sigma_star
```

Impute values based on this initial guess.

```{r impute}
Ystar <- mvnorm_fill_missing(Y, mu_star, Sigma_star)

mylegend <- function() {
  legend(
    "topleft",
    legend = c("True missing", "Present", "Imputed"),
    col = c("grey", "black", "red"),
    pch = 19,
	bg = "white"
  )
}
plot(Y_all, pch = 19, col = "grey")
points(Ystar, pch = 19, col = "red")
points(Y, pch = 19, col = "black")
mylegend()
```

Clearly, these initial values are a bad fit to the data.
Nevertheless, draw $\mu$ and $\Sigma$ conditioned on this set of imputed values.

```{r draw}
mu_star <- draw_mu(Ystar, Sigma_star, mu_0, Sigma_0)
mu_star
Sigma_star <- draw_Sigma(Ystar, mu_star, v0, S0)
Sigma_star
```

Draw a new set of imputed values, conditioned on the current $\mu^*$ and $\Sigma^*$.

```{r impute_2}
Ystar <- mvnorm_fill_missing(Y, mu_star, Sigma_star)
plot(Y_all, pch = 19, col = "grey")
points(Ystar, pch = 19, col = "red")
points(Y, pch = 19, col = "black")
mylegend()
```

This is still a bad fit to the data, but, though it is hard to tell, it has improved slightly.
Doing this repeatedly (in a loop) gradually improves the estimates of the covariance.

```{r loop}
par(mfrow = c(5, 4), mar = c(3, 3, 2, 0.1), cex = 0.4)
for (i in 1:20) {
  mu_star <- draw_mu(Ystar, Sigma_star, mu_0, Sigma_0)[1,]
  Sigma_star <- draw_Sigma(Ystar, mu_star, v0, S0)
  Ystar <- mvnorm_fill_missing(Y, mu_star, Sigma_star)
  title <- sprintf("iteration: %d, covariance: %.2f", i, Sigma_star[2, 1])
  plot(Y_all, pch = 19, col = "grey", xlab = "", ylab = "", main = title)
  points(Ystar, pch = 19, col = "red")
  points(Y, pch = 19, col = "black")
}
```

After a lot of MCMC samples, we can generate a distribution of covariance values, which provides an uncertainty estimate on the covariance.

```{r cov_estimate}
cov_ab <- numeric(1000)
for (i in seq_along(cov_ab)) {
  mu_star <- draw_mu(Ystar, Sigma_star, mu_0, Sigma_0)[1,]
  Sigma_star <- draw_Sigma(Ystar, mu_star, v0, S0)
  Ystar <- mvnorm_fill_missing(Y, mu_star, Sigma_star)
  cov_ab[i] <- Sigma_star[2, 1]
}
hist(cov_ab, xlab = "Covariance estimate", ylab = "Count", main = "Covariance estimate")
abline(v = true_cov[2, 1], col = "red", lwd = 2)
legend("topright", legend = "True cov.", lty = "solid", lwd = 2, col = "red")
```
